Ideas cascading into keystrokes -- modelling writing hesitations as finite mixture process

### 2024-06-13 report/appendix/models.Rmd
Add slides and model comparisons update

where $i$ iterates over every interkey interval $i \in 1 \dots N$ with $N$ being the total number of interkey intervals; thus $\text{location[i]}$ is the transition location associated with the $i$^\text{th} keystroke and $\text{participant[i]}$ is the participant associated with the $i$^\text{th} keystroke.

>>>>

where $i$ iterates over every interkey interval $i \in 1 \ldots N$ with $N$ being the total number of interkey intervals; thus $\text{location[i]}$ is the transition location associated with the $i$^\text{th} keystroke and $\text{participant[i]}$ is the participant associated with the $i$^\text{th} keystroke.

### 2024-06-13 report/appendix/sim.Rmd
Add slides and model comparisons update

map_dfr(files, ~read_csv(.) %>% 
        mutate(Dataset = str_remove_all(.x, "../mixture-model-sim/data|/|.csv" ))) %>%
  mutate(across(Dataset, ~recode(., mogdata = "two distributions", lmdata = "single distributions"))) %>% 
  ggplot(aes(x = value, colour = Dataset, fill = Dataset)) +
  geom_density(alpha = .25, 
               linewidth = .25) +
  scale_colour_colorblind() + 
  scale_fill_colorblind() +
  scale_x_log10(labels = ~format(., big.mark = ",")) +
  theme(text = element_text(size = 10)) +
  labs(colour = "Data process", 
       fill = "Data process", 
       x = "y", 
       y = "Density")
```

>>>>

map_dfr(files, ~read_csv(.) %>% 
        mutate(Dataset = str_remove_all(.x, "\\../|/|mixture-model-sim/data|/|\\.|csv"))) %>%
  mutate(across(Dataset, ~recode(., mogdata = "two distributions", lmdata = "single distribution"))) %>% 
  ggplot(aes(x = value, colour = Dataset, fill = Dataset)) +
  geom_density(alpha = .25, 
               linewidth = .25) +
  scale_colour_colorblind() + 
  scale_fill_colorblind() +
  scale_x_log10(labels = ~format(., big.mark = ",")) +
  theme(text = element_text(size = 10)) +
  labs(colour = "Data process", 
       fill = "Data process", 
       x = "y", 
       y = "Density")
```

### 2024-06-13 report/sections/discussion.Rmd
Add slides and model comparisons update

The classical serial view [@flower1980ynamics] characterises composing spontaneous multi-sentence text as a sequence of planning and execution cycles. Text is therefore produced as a series of production bursts bounded by pauses. The parallel model of writing, in contrast, sees  planning, at least in competent writers, as occurring to a large extent in parallel with output [@olive2014toward]. The implication of this for observation of writing timecourse -- of inter-key intervals as writers compose text -- is that pause duration doesn’t necessarily correlate with planning necessary for next output. We implemented  these two views in Bayesian statistical models, hierarchical single-distribution models capturing the serial view, and a two-distribution mixture model capturing predictions about patterning of interkey intervals under the assumption of semi-parallel upstream processing. Fitting these models with six sets of keystroke data from free text production tasks -- with a total of 953 participants producing 5,223,280 interkey intervals -- we have found compelling and consistent statistical evidence in favour of the parallel view. In all cases, the predictive performance of the two-distribution mixture-models (consistent with the parallel view of writing) outperformed the single-distribution models.

@@

 Second, pauses at before-sentence locations were not always longer than before words or mid-word; specifically there were two exceptions: (1) sentence-initial hesitations were not longer than word-initial hesitation for the C2L1 dataset possibly because these sample of young inexperienced writers tend to use a more localised word-level planning strategy for text production. (2) pause durations did not vary across transition locations in the LIFT dataset. The absence of a difference could be explained as specific to the writing task because, in synthesis, there is less need for planning novel contents and text structure: the structure of one the source texts might have been used during synthesis [“structural isomorphism”, @hidi1986producing]. However, the results of the PLanTra dataset rule out this possibility: there was evidence for longer hesitations at larger linguistic edges for a text simplification task that also does not require planning of contents.

@@

Throughout this paper we have been attributing longer (second-distribution) interkey intervals to spill-over from upstream pre-motor processing that could not be completed in its entirety in parallel with previous output. These longer inter-key intervals may also, on occasion, be associated with activity that is initiated after the preceding process and is entirely bounded by the duration of the pause. We are thinking here specifically of the tendency, particularly at higher order text boundaries, for writers to glance back into their existing text [@alamargot2010using; @beers2010adolescent; @chukharev2019combined; @de2018exploring; @torrancea; @van2010reading]. @torrance2016reading found that lookbacks appear with a frequency of 45% before sentences, 12% before words and 5% within words of which 36% were associated with sustained reading but mostly less patterned forward and backward saccades between words ["hopping", see also @chukharev2019combined]. There are broadly two reasons why writers might look back into their existing text.^[Eye movement between key presses in writing-from-sources tasks will also be associated with reading source texts. Note, however, that for  two datasets (PLanTra, LIFT) that involved writing from sources, we removed transitions that were made when the source (rather than the writing document) was in focus.]  Lookback may be associated with monitoring just-produced text for errors. In the present study we excluded interkey intervals that preceded deletion or cursor movement, and also intervals at timescales that might be associated with strategic reviewing of large spans of text (e.g., reading text-already-written from top to bottom). However it may be that some longer intervals were associated with monitoring that failed to find errors that needed correcting. A second function for lookback may be to support planning next-text [@johansson2010looking; @hayes2013new; @torrance2016reading]. The preceding text, and particularly the preceding sentence, is a rich source of cues to support retrieval of both content and lexical items necessary to prepare what to write next. This will become particularly important where upstream processing lags behind the rate at which it can be output, and one or more of the buffers necessary to resolve this “temporal friction” [@van1991handwriting] becomes overloaded. Comparison of conditions unmasked and masked text conditions in the CATO and GE2 datasets -- conditions in which writers could and could not read the text that they had already produced -- suggested that, if anything, proportion of longer interkey intervals at sentence boundaries was greater when text was masked (details in Appendix \ref{masking-effect-cato-ge2}). This finding is consistent both with lookback being only a minor contributor to interkey interval and, where it does occur, serving a rapid memory-refresh function. There is also some evidence, at least in the context of handwritten production, that eye movements back into already-written text themselves occasionally occur in parallel with output [@Alamargot2007].

The findings we report in this paper are, we argue, the first robust evidence for parallelism in written composition of multi-sentence texts. The conclusion that text production processes are  fundamentally parallel will not come as a surprise to most researchers. Parallel processing has a long history in theories of speech production [@gar75; @levelt1989speaking; @dell1986; @chang2006; @dell1999connectionist; @dell1992stages]. The transitory nature of activation and storage during fluent spoken production prevent, for example, generating phonological code a long time before it is required to drive articulation [@christiansen2016now]. Spoken production is “just-in-time”. Perhaps because, at least in principle, there is not the same push to fluency in written production -- in principle writers can pause at any point with no detriment to communication -- writing extended text has tended to be seen as fundamentally serial, with pauses to prepare next-text followed by bursts of output during which this is executed [@flower1980ynamics]. Our findings suggest that this is not the case. The language production processes that underlie text production follow the same just-in-time principle as speech.

>>>>

The classical serial view [@flower1980ynamics] characterises composing spontaneous multi-sentence text as a sequence of planning and execution cycles. Text is therefore produced as a series of production bursts bounded by pauses. The parallel model of writing, in contrast, sees  planning, at least in competent writers, as occurring to a large extent in parallel with output [@olive2014toward]. The implication of this for observation of writing timecourse -- of inter-key intervals as writers compose text -- is that pause duration doesn't necessarily correlate with planning necessary for next output. We implemented these two views in Bayesian statistical models, hierarchical single-distribution models capturing the serial view, and a two-distribution mixture model capturing predictions about patterning of interkey intervals under the assumption of semi-parallel upstream processing. Fitting these models with six sets of keystroke data from free text production tasks -- with a total of 953 participants producing 5,223,280 interkey intervals -- we have found compelling and consistent statistical evidence in favour of the parallel view. In all cases, the predictive performance of the two-distribution mixture-models (consistent with the parallel view of writing) outperformed the single-distribution models.

@@

Second, pauses at before-sentence locations were not always longer than before words or mid-word; specifically there were two exceptions: (1) sentence-initial hesitations were not longer than word-initial hesitation for the C2L1 dataset possibly because these sample of young inexperienced writers tend to use a more localised word-level planning strategy for text production. (2) pause durations did not vary across transition locations in the LIFT dataset. The absence of a difference could be explained as specific to the writing task because, in synthesis, there is less need for planning novel contents and text structure: the structure of one the source texts might have been used during synthesis ["structural isomorphism", @hidi1986producing]. However, the results of the PLanTra dataset rule out this possibility: there was evidence for longer hesitations at larger linguistic edges for a text simplification task that also does not require planning of contents.

@@

Throughout this paper we have been attributing longer (second-distribution) interkey intervals to spill-over from upstream pre-motor processing that could not be completed in its entirety in parallel with previous output. These longer inter-key intervals may also, on occasion, be associated with activity that is initiated after the preceding process and is entirely bounded by the duration of the pause. We are thinking here specifically of the tendency, particularly at higher order text boundaries, for writers to glance back into their existing text [@alamargot2010using; @beers2010adolescent; @chukharev2019combined; @de2018exploring; @torrancea; @van2010reading]. @torrance2016reading found that lookbacks appear with a frequency of 45% before sentences, 12% before words and 5% within words of which 36% were associated with sustained reading but mostly less patterned forward and backward saccades between words ["hopping", see also @chukharev2019combined]. There are broadly two reasons why writers might look back into their existing text.^[Eye movement between key presses in writing-from-sources tasks will also be associated with reading source texts. Note, however, that for two datasets (PLanTra, LIFT) that involved writing from sources, we removed transitions that were made when the source (rather than the writing document) was in focus.] Lookback may be associated with monitoring just-produced text for errors. In the present study we excluded interkey intervals that preceded deletion or cursor movement, and also intervals at timescales that might be associated with strategic reviewing of large spans of text (e.g., reading text-already-written from top to bottom). However it may be that some longer intervals were associated with monitoring that failed to find errors that needed correcting. A second function for lookback may be to support planning next-text [@johansson2010looking; @hayes2013new; @torrance2016reading]. The preceding text, and particularly the preceding sentence, is a rich source of cues to support retrieval of both content and lexical items necessary to prepare what to write next. This will become particularly important where upstream processing lags behind the rate at which it can be output, and one or more of the buffers necessary to resolve this "temporal friction" [@van1991handwriting] becomes overloaded. Comparison of conditions unmasked and masked text conditions in the CATO and GE2 datasets -- conditions in which writers could and could not read the text that they had already produced -- suggested that, if anything, proportion of longer interkey intervals at sentence boundaries was greater when text was masked (details in Appendix \ref{masking-effect-cato-ge2}). This finding is consistent both with lookback being only a minor contributor to interkey interval and, where it does occur, serving a rapid memory-refresh function. There is also some evidence, at least in the context of handwritten production, that eye movements back into already-written text themselves occasionally occur in parallel with output [@Alamargot2007].

The findings we report in this paper are, we argue, the first robust evidence for parallelism in written composition of multi-sentence texts. The conclusion that text production processes are  fundamentally parallel will not come as a surprise to most researchers. Parallel processing has a long history in theories of speech production [@gar75; @levelt1989speaking; @dell1986; @chang2006; @dell1999connectionist; @dell1992stages]. The transitory nature of activation and storage during fluent spoken production prevent, for example, generating phonological code a long time before it is required to drive articulation [@christiansen2016now]. Spoken production is "just-in-time". Perhaps because, at least in principle, there is not the same push to fluency in written production -- in principle writers can pause at any point with no detriment to communication -- writing extended text has tended to be seen as fundamentally serial, with pauses to prepare next-text followed by bursts of output during which this is executed [@flower1980ynamics]. Our findings suggest that this is not the case. The language production processes that underlie text production follow the same just-in-time principle as speech.

### 2024-06-13 report/sections/introduction.Rmd
Add slides and model comparisons update

The coordination of these processes in writing is arguably more demanding than in spoken language. This is because writing requires spelling retrieval, which may or may not be mediated by phonology, and the motor planning and execution required to press keys or move a pen. Pen control and typing skills are later-developing and more effortful than those required to articulate language sounds in speech and, possibly as a consequence, written output rate is typically slower [see for a recent review @kandel2023written]. Consequently mentally buffering of linguistic plans prior to output is likely to play a greater role in writing than in speech.

@@

Capturing this understanding of interkey intervals requires statistical models that are premised on more than one data-generating process. These models have been adopted in a number of psycholinguistic studies. For example, in language comprehension, Vasishth and colleagues [@vasishth2017feature; @vasishth2017] demonstrated convincingly that some phenomena in online sentence comprehension observed in reading times are captured better by statistical models that assume that the underlying mental process that generates reading data is in fact a mixture of two process expressed as two distributions in the statistical implemental. In the context of written production @roeser2021modelling demonstrated^[For similar approaches see  @almond2012preliminary, @baaijen2012keystroke, @chenu2014interword, @guo2018modeling, @hall2022constructing, @li2021identifying, and @waes2019.]  how models that allow the possibility of two underlying data-generating processes revealed two separate distributions when participants were copy-typing sentences or lists of words that are more pronounced for tasks with challenging orthographic sequences but reduced down to just one distribution in a tapping task where participants merely pressed two keys repeatedly as quickly as possible (e.g. “dkdkdkdk”). As argued by the authors, copy-typing underlies a similar cascade of processes as free text composition: copy-typing provides a typing context that is constrained to visual encoding of the target string and the retrieval of motor codes for letter combinations. For relatively experienced writers, visual encoding is likely to happen in parallel to writing [@de2018exploring]; hesitations occur when motor codes, for example for low frequency bigrams, were not retrieved in time [@roeser2021modelling]. However composition tasks are different from copy tasks: in copy tasks but not in spontaneous text production, content generation and grammatical encoding is artificially constrained to particular sequences of letter bigrams, words or sentences, and copy-typing involves a combination of visual encoding of the target string and mental buffering prior to the activation of motor codes. Typing hesitations in the context of copy-typing may – to some extent – reflect not just production but also, and possibly largely, visual encoding. Text composition, in contrast, provides data from the free generation of words and full sentences that is not dependent on a visually presented text stimulus. For these reasons our research extends the analysis of keystroke data using a mixture model approach to data from spontaneous text production.

>>>>

The coordination of these processes in writing is arguably more demanding than in spoken language. This is because writing requires spelling retrieval, which may or may not be mediated by phonology, and the motor planning and execution required to press keys or move a pen. Pen control and typing skills are later-developing and more effortful than those required to articulate language sounds in speech and, possibly as a consequence, written output rate is typically slower [for a recent review see @kandel2023written]. Consequently mentally buffering of linguistic plans prior to output is likely to play a greater role in writing than in speech.

@@

Capturing this understanding of interkey intervals requires statistical models that are premised on more than one data-generating process. These models have been adopted in a number of psycholinguistic studies. For example, in language comprehension, Vasishth and colleagues [@vasishth2017feature; @vasishth2017] demonstrated convincingly that some phenomena in online sentence comprehension observed in reading times are captured better by statistical models that assume that the underlying mental process that generates reading data is in fact a mixture of two process expressed as two distributions in the statistical implementation. In the context of written production @roeser2021modelling demonstrated^[For similar approaches see  @almond2012preliminary, @baaijen2012keystroke, @chenu2014interword, @guo2018modeling, @hall2022constructing, @li2021identifying, and @waes2019.] how models that allow the possibility of two underlying data-generating processes revealed two separate distributions when participants were copy-typing sentences or lists of words that are more pronounced for tasks with challenging orthographic sequences but reduced down to just one distribution in a tapping task where participants merely pressed two keys repeatedly as quickly as possible (e.g. “dkdkdkdk”). As argued by the authors, copy-typing underlies a similar cascade of processes as free text composition: copy-typing provides a typing context that is constrained to visual encoding of the target string and the retrieval of motor codes for letter combinations. For relatively experienced writers, visual encoding is likely to happen in parallel to writing [@de2018exploring]; hesitations occur when motor codes, for example for low frequency bigrams, were not retrieved in time [@roeser2021modelling]. However composition tasks are different from copy tasks: in copy tasks but not in spontaneous text production, content generation and grammatical encoding is artificially constrained to particular sequences of letter bigrams, words or sentences, and copy-typing involves a combination of visual encoding of the target string and mental buffering prior to the activation of motor codes. Typing hesitations in the context of copy-typing may – to some extent – reflect not just production but also, and possibly largely, visual encoding. Text composition, in contrast, provides data from the free generation of words and full sentences that is not dependent on a visually presented text stimulus. For these reasons our research extends the analysis of keystroke data using a mixture model approach to data from spontaneous text production.

### 2024-06-13 report/sections/methods.Rmd
Add slides and model comparisons update

The first line of equation \ref{eq:bimodcon2} represents the distribution of hesitant interkey intervals that include the $\delta$ parameter presenting the slowdown for hesitant transitions; the second line represents the distribution of fluent -- uninterrupted -- interkey intervals that is characterised by $\beta$ which is also present in the first file. Either of these two distributions is associated with the mixing proportion $\theta$ that ranges between 0 and 1. $\theta$ was parameterised here to represent the probability that an interkey interval is associated with the distribution of hesitant transitions. This probability is inversely related to the mixing weight of the distribution of fluent transitions by $1-\theta$ . In other words, a larger weight for either distribution inevitably means a lower weight for the other. We call this parameter the probability of hesitant interkey intervals. When a parameter is allowed to vary by transition location (levels: before sentence, before word, within word), this was indicated as subscript; similar for participants. For example, $\theta_\text{location[i], participant[i]}$ means that the mixing proportion $\theta$ is allowed to take on a different value for both the transition location and participant associated with the $i^\text{th}$ keystroke interval $i \dots \in N$ where $N$ is the total number of keystroke intervals. Because the hesitation probability was allowed to vary by participants as well as location type, we included a hyper-parameter for the mixing proportion $\theta$ for each transition location.

>>>>

The first line of equation \ref{eq:bimodcon2} represents the distribution of hesitant interkey intervals that include the $\delta$ parameter presenting the slowdown for hesitant transitions; the second line represents the distribution of fluent -- uninterrupted -- interkey intervals that is characterised by $\beta$ which is also present in the first file. Either of these two distributions is associated with the mixing proportion $\theta$ that ranges between 0 and 1. $\theta$ was parameterised here to represent the probability that an interkey interval is associated with the distribution of hesitant transitions. This probability is inversely related to the mixing weight of the distribution of fluent transitions by $1-\theta$ . In other words, a larger weight for either distribution inevitably means a lower weight for the other. We call this parameter the probability of hesitant interkey intervals. When a parameter is allowed to vary by transition location (levels: before sentence, before word, within word), this was indicated as subscript; similar for participants. For example, $\theta_\text{location[i], participant[i]}$ means that the mixing proportion $\theta$ is allowed to take on a different value for both the transition location and participant associated with the $i^\text{th}$ keystroke interval $i \in 1 \ldots N$ where $N$ is the total number of keystroke intervals. Because the hesitation probability was allowed to vary by participants as well as location type, we included a hyper-parameter for the mixing proportion $\theta$ for each transition location.

### 2024-06-13 report/sections/results.Rmd
Add slides and model comparisons update

```{r}
files <- list.files(
  str_c("../../stanout/", 
        c("lift", "spl2", "spl2_shift", "plantra", "cato", "c2l1", "gunnexp2")), 
  pattern = "model.+.csv", 
  full.names = T)

>>>>

```{r message=FALSE}
files <- list.files(
  str_c("../../stanout/", 
        c("lift", "spl2", "spl2_shift", "plantra", "cato", "c2l1", "gunnexp2")), 
  pattern = "model.+.csv", 
  full.names = T)
